---
title: "Biostat HW3"
author: "Alyssa Vanderbeek"
date: "12 October 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library(tidyverse)
```

### Problem 3: A rehabilitation center is interested in examining the relationship between physical status before therapy and the time (days) required in physical therapy until successful rehabilitation. Records from patients 18-30 years old were collected and provided to you for statistical analysis (data “Knee.csv”).

### 1. Generate descriptive statistics for each group and comment on the differences observed

```{r}
# import data
knee_data = read_csv('~/Desktop/FALL 2018/Biostat I/Homework/data/HW3_export/Knee.csv', col_types = cols()) 

# produce table of summary statistics 
apply(X = knee_data, FUN = summary, MARGIN = 2) %>% # apply the summary function to each column of the data
  do.call(bind_rows, .) %>% # take list produced by apply and turn into tibble
  mutate('Physical Health' = colnames(knee_data)) %>% # make a column specifying what group each row corresponds to
  select(`Physical Health`, everything()) %>% # rearrange column order
  knitr::kable(caption = 'Summary statistics across') # make nice-looking table

# boxplot of distribution of days across physical health groups
knee_data = knee_data %>%
  gather(key = 'physical_status', value = 'days') 

knee_data %>% # long format for easy plotting
  ggplot(aes(x = physical_status, y = days)) +
    geom_boxplot() +
    labs(title = 'Number of days in physical therapy',
         x = 'Physical health status',
         y = 'Days') 

# ANOVA table
anova(lm(days~factor(physical_status), data = knee_data)) %>%
  knitr::kable()

# multiple testing adjustments
pairwise.t.test(knee_data$days, knee_data$physical_status, p.adj = 'bonferroni')

```

